---
title: "Analysis script"
author: "Sunyoung Park"
date: "03/31/2023"
output: html_document
---

  
This Quarto file loads the cleaned Palmer Penguins data and does some initial analysis. 


# Analysis 

In this analysis script, I am analyzing and answering four questions based on the Palmer Penguins dataset. The four questions include:

* Do species differ in culmen length?
* Does sex have an effect on culmen length?
* Does Island have an effect on culmen length?
* Do size and species have an effect on culmen length?

# Setup

```{r }
#| cache = FALSE
knitr::read_chunk('statistical_analysis.R')
```

Load needed packages and filepaths.  

```{r setup}
#| message: FALSE
#| warning: FALSE
```

Load custom functions.

```{r functions}
```

Load the data.

```{r loaddata}
```

# Summary table

Summary table consists of basic statistics for the data, including sample size, counts, and means and standard errors for continuous data.

Here we will use `skimr` which produces a summary table, but with more than we want. Letʻs save it as an object so we can grab the elements we want. 

It turns out that the `skim()` function returns a tibble. We can coerce it to a dataframe and use `head()` to see what the column names are and what the data look like. 

```{r, summarize}
```

## Select columns for a final summary table

Letʻs retain only the variable, N, mean, sd, and category counts, but letʻs rename them. Since we will use the SD to compute the standard error, letʻs just rename it as SE now and save a step later.  (Remember the standard error is just the standard deviation/ square root(sample size)). 

Note: The `results="asis"` is an option for `knitr` to layout the the results without the code formatting. Normally output from code chunks has the look of R output. `"asis"` removes that behavior, so we can lay it out as a typeset table for print. 

```{r summary.table, results="asis"}
```

# Results of the data analysis

The results were created by the `statistical_analysis.R` script and to the `Results` folder. There are four subfolders in the `Results` folder named Q1, Q2, Q3 and Q4, which correspond to the four questions that will follow below, respectively.


# 1. Do species differ in culmen length?

First, we begin by plotting a boxplot of culmen length by species:

```{r culmenL_species_boxplot}
```

The boxplot shows that individuals of Adelie species have the shortest culmen length, while Chinstrap has the longest, followed by Gentoo which is slightly shorter than Chinstrap. 


Now letʻs try a density plot

```{r culmenL_species_density}
```

The density plot also demonstrates similar results the boxplot above; Adelie has the highest density in the shortest culmen length, and Chinstrap shows the highest density in the longest culmen length. 

To test the statistical significance of this result, we can use ANOVA. 

We see that species do differ in culmen length and this is a very robust statistical result; p-value is <0.001, with a f-value of >400.  

```{r culmenL_species_aov}
```

# 2. Does sex have an effect on culmen length?


What about differences in culmen length affected by sex? Since we know that species differ, we can include a two-way ANOVA and test both factors. 


When we add sex to the model _after_ we account for species, both species and sex are statistically significant, where species is showing a higher variation and more significance. 

```{r culmenL_species_sex_aov}
```

If we add sex _first, before_ species, it is still significant. We interpret this as sex differing in culmen length.

```{r culmenL_sex_species_aov}
```


Letʻs plot both factors together to see whatʻs going on. 

```{r culmenL_species_sex_density }
```

Males for all three species seem to have longer culmen lengths compared to females. Within each plot, the density between the three species are quite different with Adelie having the shortest culmen length and being the least overlapping with the other two species.

Two-way ANOVA tests show that these above results are statistically significant; sex has an effect on culmen length. 



# 3. Does Island have an effect on culmen length?


What about differences in culmen length affected by island? Since we know that species differ, we can include a two-way ANOVA and test both factors, similar to Q2 above.  


When we add island to the model _after_ we account for species, it is not significant (p-value >0.05).

```{r culmenL_species_island_aov}
```

If we add island _first, before_ species, it is significant (p-value <0.05). We interpret this as island differing in culmen length.

```{r culmenL_island_species_aov}
```


Letʻs plot both factors together to see whatʻs going on. 

```{r culmenL_species_island_density }
```


# 4. Do size and species have an effect on culmen length?

Now we want to know if body mass and species have an effect on culmen length. 



First, we plot both factors. 

From the plot, it appears that body mass and culmen length are positively related in all three species. Chinstrap has longer culmen length with smaller body mass, Gentoo has longer culmen length with higher body mass, and Adelie has shorter culmen length with smaller body mass.

```{r culmenL_mass_species_plot }
```


Now we use ANCOVA model using culmen length (continuous response variable) as outcome, and species (categorical explanatory variable) and body mass (continuous explanatory variable) as predictors. This model is used to assess whether the examined differences in the plot above are statistically significant or could be expected by chance. 


```{r culmenL_mass_species_aov }
```


This model assumes that the two explanatory factors have _interaction_, meaning whether the covariate, body mass, depends on the categorical predictor, species. 

The p-value for species is <0.05 and therefore, culmen length varies across all three species.

The p-value for body mass is also <0.05, meaning it is statistically significant that culmen length increases with body mass. 

However, Pr(>F) value for _body mass:species_ is >0.05, and therefore, the statement "the covariate, body mass, depends on the categorical predictor, species" is not statistically significant, which means that the variation observed in the plot would very likely happen "by chance." 



